A matrix-free approach for finite-strain hyperelastic problems using geometric multigrid

### 2018-10-24 doc/large_strain.tex
minor update of fig title
update Linearization section based on discussion with Steinmann

The principle of stationary potential energy at equilibrium requires that the directional derivative with respect to the displacement
\begin{align}
\delta \mcl E =
\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, .
\label{eq:stationary}
\end{align}
vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
%
This leads to the following scalar non-linear equation
\begin{align}
F(\gz u, \delta \gz u) =
\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \, \d V \,
-
\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
= 0 \, ,
\end{align}
where $\gz P := \partial \mcl \psi / \partial \gz F$ is the Piola stress tensor.
The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
\begin{align}
\gz P : \Grad \, \delta \gz u =
\left[\gz \tau \cdot \gz F^{-T}\right] : \Grad \, \delta \gz u =
\gz \tau : \left[\Grad \, \delta \gz u \cdot \gz F^{-1}\right] =
\gz \tau : \grad \, \delta \gz u \, ,
\end{align}
and therefore
\begin{align}
  F(\gz u, \delta \gz u) =
  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
  = 0 \, ,
\label{eq:weak_form}
\end{align}
where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.

@@

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u.
\end{align}
For conservative traction boundary conditions, the directional derivative is given by
\begin{equation}
\begin{split}
\D F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
&=
\int_{\mcl B_0} \left[\D \left[\gz F \cdot \gz S \cdot \gz F^T\right] \cdot \Delta \gz u\right] : \overline{\grad^s} \, \delta \gz u \, \d V
\\
& +
\int_{\mcl B_0} \overline{\gz \tau} : \left[ \Grad \, \delta \gz u \cdot
\left[\D \gz F^{-1} \cdot \Delta \gz u\right]
\right] \d V.
\end{split}
\end{equation}
Following \cite{Wriggers2008}, this can be simplified to
\begin{equation}
  \begin{split}
\D F(\overline{\gz u}, \delta \gz u) \cdot \Delta \gz u
  &=
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V \\
  &+
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\end{split}
\label{eq:tangent}
\end{equation}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and therefore called material part of the tangent operator.  The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{coarse_mesh.png}
      \caption{2D coarse mesh}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{coarse_mesh_3d.png}
    \caption{3D coarse mesh}
  \end{subfigure}
  \caption{Discretization of the heterogeneous material at the coarsest mesh level.}%
  \label{fig:miehe}
\end{figure}

>>>>

The principle of stationary potential energy at equilibrium requires that the directional derivative with respect to the displacement
\begin{align}
\D_{\displaystyle \delta \gz u} \mcl E :=
\frac{\d}{\d \epsilon} \mcl E (\gz u + \epsilon \delta \gz u) \Bigr\rvert_{\epsilon=0}  = 0 \, \qquad \forall \delta \gz u \, .
\label{eq:stationary}
\end{align}
vanishes for all directions $\delta \gz u$ which satisfy homogeneous Dirichlet boundary conditions.
%
This leads to the following scalar non-linear equation
\begin{align}
F(\gz u, \delta \gz u) =
\int_{\mcl B_0} \gz P : \Grad \, \delta \gz u \, \d V \,
-
\int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
= 0 \, ,
\end{align}
where $\gz P := \partial \mcl \psi / \partial \gz F$ is the Piola stress tensor.
The double contraction in the first term can be re-written in terms of the symmetric Kirchhoff stress tensor $\gz \tau := \gz P \cdot \gz F^T$ as
\begin{align}
\gz P : \Grad \, \delta \gz u =
\left[\gz \tau \cdot \gz F^{-T}\right] : \Grad \, \delta \gz u =
\gz \tau : \left[\Grad \, \delta \gz u \cdot \gz F^{-1}\right] =
\gz \tau : \grad \, \delta \gz u \, ,
\end{align}
and therefore
\begin{align}
  F(\gz u, \delta \gz u) =
  \int_{\mcl B_0} \gz \tau : \grad^{s} \, \delta \gz u \, \d V \,
  -
  \int_{\partial \mcl B_0^N} \overline{\gz T} \cdot \delta \gz u \, \d S
  = 0 \, ,
\label{eq:weak_form}
\end{align}
where $\grad^{s} \, \delta \gz u := \frac{1}{2}\left[ \grad \, \delta \gz u + ( \grad \, \delta \gz u)^T\right]$.
Note that $\gz \tau$ is the push-forward transformation of the Piola-Kirchhoff stress $\gz S := \partial \mcl \psi / \partial \gz E \equiv 2 \partial \mcl \psi / \partial \gz C$,
that is $\gz \tau = \chi(\gz S) = \gz F \cdot \gz S \cdot \gz F^T$.

@@

In order to solve \eqref{eq:weak_form} using Newton's method, a first order approximation around a given solution field $\overline{\gz u}$ is required such that
\begin{align}
F(\overline{\gz u} + \Delta \gz u, \delta \gz u) \approx
F(\overline{\gz u}, \delta \gz u) + \D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u) ,
\end{align}
where $\D_{\displaystyle \Delta \gz u}(\bullet)$ denotes the directional derivative in the direction $\Delta \gz u$.
For conservative traction boundary conditions, the directional derivative is given by
\begin{equation}
\begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
&=
\int_{\mcl B_0}
\D_{\displaystyle \Delta \gz u} \left(\gz F \cdot \gz S \cdot \gz F^T\right)  :
\overline{\grad^s} \, \delta \gz u
\, \d V
\\
& +
\int_{\mcl B_0}
\overline{\gz \tau} :
\left[
  \Grad \, \delta \gz u \cdot
  \D_{\displaystyle \Delta \gz u} \gz F^{-1}
\right] \d V.
\end{split}
\label{eq:tangent_pre}
\end{equation}
Following \cite{Wriggers2008}, this can be simplified to\footnote{To that end
$\D_{\displaystyle \Delta \gz u} \gz F = \Grad \Delta \gz u$,
$\;\D_{\displaystyle \Delta \gz u} \gz F^{-1} = - \gz F^{-1} \cdot \D_{\displaystyle \Delta \gz u} \gz F \cdot \gz F^{-1}$ and
$\D_{\displaystyle \Delta \gz u} \gz E = \frac{1}{2}\left[\D_{\displaystyle \Delta \gz u} \gz F^T \cdot \gz F + \gz F^T \cdot \D_{\displaystyle \Delta \gz u} \gz F\right]$
are employed together with
$\D_{\displaystyle \Delta \gz u} \gz S = 2 \partial \gz S / \partial \gz C : \D_{\displaystyle \Delta \gz u} \gz E$.
}
\begin{equation}
  \begin{split}
\D_{\displaystyle \Delta \gz u} F(\overline{\gz u}, \delta \gz u)
  &=
  \int_{\mcl B_0} \overline{\grad^s} \Delta \gz u : J \boldsymbol{\mathcal{C}} : \overline{\grad^s} \, \delta \gz u \, \d V \\
  &+
  \int_{\mcl B_0}
  \overline{\grad}\delta \gz u :
  \left[
  \overline{\grad} \Delta \gz u \cdot
  \overline{\gz \tau}
  \right]
  \d V.
\end{split}
\label{eq:tangent}
\end{equation}
Here $\overline{(\bullet)}$ is used to denote quantities evaluated using the displacement field $\overline {\gz u}$, and the fourth-order tangent tensor is the push forward of the material tangent $J \boldsymbol{\mathcal{C}} = \chi\left( 4 \frac{d^{2} \psi \left( \mathbf{C} \right)}{d \mathbf{C} \otimes d \mathbf{C}} \right)$. The first term in \eqref{eq:tangent} is related to the linearization of the Piola-Kirchhoff stress $\gz S$ and therefore called material part of the tangent operator.  The second term in \eqref{eq:tangent} is called geometric part of the tangent since it originates from the linearization of $\gz F$, $\gz F^T$ and $\gz F^{-1}$.

@@

\begin{figure}[!ht]
  \centering
  \begin{subfigure}[b]{0.45\textwidth}
      \centering
      \includegraphics[width=\textwidth]{coarse_mesh.png}
      \caption{2D coarse mesh}
  \end{subfigure}
  \begin{subfigure}[b]{0.45\textwidth}
    \centering
    \includegraphics[width=\textwidth]{coarse_mesh_3d.png}
    \caption{3D coarse mesh}
  \end{subfigure}
  \caption{Discretization of the heterogeneous material at the coarsest mesh level and the prescribed boundary conditions.}%
  \label{fig:miehe}
\end{figure}
